AusTraits -- a curated plant trait database for the Australian flora

### 2020-10-28 ms.Rmd
Rework: Adding Description file, combine sources, misc changes

---
title: "Austraits -- a curated plant trait database for the Australian flora"
csl: assets/nature.csl
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
  word_document:
    reference_docx: assets/manuscriptstyle.docx
header-includes:
- \usepackage{inputenc}
- \usepackage[T1]{fontenc}
- \usepackage{float}
- \usepackage{fontspec}
- \usepackage{unicode-math}
# - \setmainfont{Arial}
# - \setmathfont{Cambria Math}
bibliography: [data/references_ms.bib]
---

@@

austraits <- readRDS("data/austraits_2.0.0.rds")
```

@@

data_contributors <- 
  austraits$contributors %>% 
    filter(
          # old studies
          !dataset_id %in% c("Lawes_2012", "Westman_1977", "Hall_1981", "Hocking_1982", "Hocking_1986", "Toelken_1996", "Brock_1993", "Forster_1992", "Forster_1995", "Goble_1981"),
           !role %in% c("assistant"),
           # deceased
           !name %in% c("Peter Clarke", "Barbara Rice", "Peter Myerscough", "Lyn Craven", "Harold Trevor Clifford", "William Cooper")
           ) %>% 
    mutate(year = dataset_id %>% str_split_fixed("_", 4) %>% .[,2]) %>% 
    arrange(year) %>%
    # take most recent address for each author
    group_by(name) %>% 
    summarise(institution = last(institution),
              dataset_ids = paste(dataset_id, collapse = ", ")) %>% 
    ungroup() %>% 
    distinct() %>% 
    mutate(surname = str_split(name, " ") %>% map_chr(last)) %>% 
    rename(affiliation = institution) %>%
#    bind_rows(
#      read_csv("data/data_pending.csv")) %>%
    arrange(surname) %>% 
    select(-surname)

@@

#  authors %>% select(dataset_ids, everything()) %>% mutate_all(~ replace_na(.x, "")) %>% write_csv("data/author_list.csv")

@@

## Todo

Refs

* Australian Plant Census, Centre of Australian National Biodiversity Research, Council of Heads of Australasian Herbaria, (14/05/2020) https://id.biodiversity.org.au/tree/51354547
* Australian Plant Name Index (continuously updated), Centre of Australian National Biodiversity Research, www.biodiversity.org.au/nsl/services/apni (14/05/2020)

@@

@@

AusTraits was assembled from `r austraits$sources %>% length()` distinct sources, including published literature, field campaigns, floras, and taxonomic treatments. We identified a list of candidate traits of interest, including growth form, plant height, leaf size, specific leaf area and seed mass, though several other traits were also collected opportunistically. A full list of traits and their sources appears in Table 1 (available online only). We then identified primary sources containing measurements for these traits and sought access to this data, including both data and metadata. Each original source was designated a `dataset_id`. Ordinarily the id is the surname of first author and year of publication, for the source’s primary citation (e.g. Falster_2005, for Dataset citation 40). Details from each source were saved with minimal modification into two plain text files:

@@

@@

@@

To harmonise each source into the common AusTraits format we applied a reproducible and transparent workflow (Figure 1), written in R [@R-2020] and using custom code and drawing on functions from the packages `tidyverse` [@wickham2016tidyverse], `stringr` [@wickham2017stringr], `yaml` [@stephens2014yaml], `remake` [@fitzjohn2016remake],  `knitr` [@xie1general] and `rmarkdown` [@allaire2015rmarkdown]. In this workflow, we performed a series of operations, including reformatting data into standardising format, generating observation id’s for each individual measured, transforming variable names into common terms, transforming data common units, standardising terms for categorial variables, encoding suitable metadata, and flagging data that did not pass quality checks.

@@

@@

DSF was supported by Australian Research Council Future Fellowship (FT160100113), RVG was supported by Australian Research Council DECRA Fellowship (DE170100208).

We acknowledge XXXXXXXXXXx institutions, who

@@

RVG, IJW conceived the original idea; RVG, EHW, CB collated data from primary sources; DSF developed the workflow for the harmonising of data and led all coding; EHW, DI, SCA, JL contributed to coding; EHW, CB, JL, SA error-checked trait observations; DI developed figures for the paper; DSF, RVG, DI, EHW wrote the first draft of the paper. All other authors contributed the raw data and metadata undeprinning the resource and reviewed the harmonised data for errors.

@@

```{r austraits}
## Assumes to items exist in global name space
austraits <- readRDS("data/austraits_2.0.0.rds")

definitions <- austraits$definitions

tmp <- RefManageR::WriteBib(austraits$sources, "data/references_data.bib")

@@

sources_all <- 
  bind_rows(
    austraits$methods %>% select(dataset_id, citation = source_primary_key) %>% 
            distinct() %>% mutate(type="primary"),
    austraits$methods %>% select(dataset_id, citation = source_secondary_key) %>% 
            distinct() %>% mutate(type="secondary")
  ) %>% 
  na.omit() %>% 
  arrange(dataset_id) %>% 
  group_by(dataset_id) %>% 
  summarise(refs = sprintf("@%s", citation) %>% 
  paste(collapse = "; "))

@@

<!-- Cite: -->
  <!-- G Yu, DK Smith, H Zhu, Y Guan, TTY Lam*. ggtree: an R package for visualization and annotation of phylogenetic trees with their covariates and other associated data. Methods in Ecology and Evolution. 2017, 8(1):28-36. doi: 10.1111/2041-210X.12628.  -->
  <!-- G Yu*, TTY Lam, H Zhu, Y Guan*. Two methods for mapping and visualizing associated data on phylogeny using ggtree. Molecular Biology and Evolution, 2018, 35(2):3041-3043. doi: 10.1093/molbev/msy194. -->

@@

@@

n_records <- 
  data_geo %>% 
  group_by(trait_name) %>%
  summarise(
            Description = "",
            Type= "",
            `Records (all)` = n(),
            `Records (geo.)` = sum(!is.na(`longitude (deg)`) & !is.na(`latitude (deg)`)),
            Studies = n_distinct(dataset_id),
            Taxon = n_distinct(taxon_name),
            Families = n_distinct(family),
            `Data Citations` = refs %>% unique() %>% paste(collapse = "; ")
            ) %>% 
  left_join(trait_categories, by=c("trait_name")) %>% 
  select(Tissue = tissue, Category = category, Trait = trait_name, everything()) %>% 
  arrange(Tissue, Category, Trait)

n_records$Type <- sapply(austraits$definitions$traits$elements, "[[","type")[n_records$Trait]
n_records$Description <- sapply(austraits$definitions$traits$elements, "[[","description")[n_records$Trait]

n_records %>%
  select(-Tissue, Description) %>%
  kable()
```

@@

### Data sources:

#### Version 1 - via bib.print

```{r, results='asis', eval=TRUE}
sources_unique <-
  bind_rows(
    austraits$methods %>% select(citation = source_primary_citation)  %>% mutate(type="primary"),
    austraits$methods %>% select(dataset_id, citation = source_secondary_citation) %>% mutate(type="secondary")
  ) %>%
  select(citation) %>%
            distinct() %>% 
            mutate(source_number=1:n())

sprintf("%s. %s", sources_unique$source_number, sources_unique$citation) %>% paste(collapse = "\n\n") %>% writeLines()

#### Version 3 - via csl

```{r, results='asis', eval=FALSE}

austraits$sources %>% sapply(function(i) i$key) -> keys

sprintf("@%s",keys) %>% paste(collapse = "; ") %>% writeLines()
```

>>>>

---
title: "AusTraits -- a curated plant trait database for the Australian flora"
csl: assets/nature.csl
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
  word_document:
    reference_docx: assets/manuscriptstyle.docx
header-includes:
- \usepackage{inputenc}
- \usepackage[T1]{fontenc}
- \usepackage{float}
- \usepackage{fontspec}
- \usepackage{unicode-math}
# - \setmainfont{Arial}
# - \setmathfont{Cambria Math}
bibliography: [data/references_all.bib]
---

@@

austraits <- readRDS("data/austraits_2.0.0.rds")

sources_all <- 
  bibtex::read.bib("data/references_ms.bib")

RefManageR::WriteBib(austraits$sources, "data/references_data.bib")

c(readLines("data/references_ms.bib"), "\n", readLines("data/references_data.bib")) %>% writeLines("data/references_all.bib")

@@

# check list of acknowledgements and assistants

data_contributors <- 
  austraits$contributors %>% 
    filter(
          # old studies or excluded
          !dataset_id %in% c("Lawes_2012", "Westman_1977", "Hall_1981", "Hocking_1982", "Hocking_1986", "Toelken_1996", "Brock_1993", "Forster_1992", "Forster_1995", "Goble_1981"),
           !role %in% c("assistant"),
           # deceased
           !name %in% c("Peter Clarke", "Barbara Rice", "Peter Myerscough", "Lyn Craven", "Harold Trevor Clifford", "William Cooper"),
            # no response or can't locate
           !name %in% c("Huw Morgan", "Wendy Cooper", "Muhammad Islam", "Michael Lawes")
           ) %>% 
    mutate(year = dataset_id %>% str_split_fixed("_", 4) %>% .[,2]) %>% 
    arrange(year) %>%
    # take most recent address for each author
    group_by(name) %>% 
    summarise(institution = last(institution),
              dataset_ids = paste(dataset_id, collapse = ", ")) %>% 
    ungroup() %>% 
    distinct() %>% 
    mutate(surname = str_split(name, " ") %>% map_chr(last)) %>% 
    rename(affiliation = institution) %>%
#    bind_rows(
#      read_csv("data/data_pending.csv")) %>%
    arrange(surname) %>% 
    select(-surname)

@@

authors %>% select(dataset_ids, everything()) %>% mutate_all(~ replace_na(.x, "")) %>% write_csv("data/author_list.csv")

@@

@@

## Comparison to TRY

numbers of taxa / figures

* AusTraits includes more seed and fruit nutrient data; possibly reflecting the interest in Australia, specifically WA, in understanding how fruit and seeds are provisioned in nutrient depauperate environments
* AusTraits includes more categorical values, especially variables documenting different components of species’ fire response strategies, reflecting the importance of fire in shaping Australian communities and the research to document different strategies species have evolved to succeed in fire-prone environments

- On the point about many Australian species being unrepresented in global trait databases (end of the second paragraph in Background & Summary) -- I was wondering if this could be exemplified by reporting the number of species in AusTraits compared to the number of Australian species in the TRY Plant Database (in order to show that AusTraits fills this gap substantially).
- I was also wondering if the number of studies that have used global plant trait databases could be reported as an example to show how valuable AusTraits will be (e.g. TRY Plant Trait Database having 1613 citations according to Google Scholar) (relevant to the point that is made in the second-last sentence of Background & Summary).

@@

AusTraits was assembled from `r austraits$sources %>% length()` distinct sources, including published papers, field campaigns, botanical collections, and taxonomic treatments. Initially we identified a list of candidate traits of interest -- including growth form, plant height, leaf size, specific leaf area and seed mass -- then identified primary sources containing measurements for these trait, before contacting authors for access. As the compilation grew, we expanded the list of traits considered to include any measureable quantity that had been quantified for a moderate number of taxa ( n > 20). A full list of traits and their sources appears in Table 1 (available online only). Details from each source were saved with minimal modification into two plain text files:

@@

```{r, results="asis"}
austraits$sources %>% sapply(function(i) i$key) -> keys

sprintf("@%s",keys) %>% paste(collapse = "; ") %>% paste0("[", ., "]") %>% writeLines()
```

### Trait definitions

TOP: [Garnier-2017]

### Taxonomy

Taxonomy updated via Australian Plant Census (APC) [@apc-2020], Australian Plant Names Index (APNI) [@apni-2020]

@@

Each original source was designated a `dataset_id`. Ordinarily the id is the surname of first author and year of publication, for the source’s primary citation (e.g. Falster_2005, for Dataset citation 40).

@@

To harmonise each source into the common AusTraits format we applied a reproducible and transparent workflow (Figure 1), written in R [@R-2020] and using custom code and drawing on functions from the packages `tidyverse` [@wickham2016tidyverse], `stringr` [@wickham2017stringr], `yaml` [@stephens2014yaml], `remake` [@fitzjohn2016remake],  `knitr` [@xie1general] and `rmarkdown` [@allaire2015rmarkdown]. In this workflow, we performed a series of operations, including reformatting data into standardising format, generating observation id’s for each individual measured, transforming variable names into common terms, transforming data common units, standardising terms for categorical variables, encoding suitable metadata, and flagging data that did not pass quality checks.

@@

#### Analysis

Phylogenetic tree constructed by pruning  [@Smith-2018] using package `V.PhyloMaker` [@Jin-2020] and visualising using the package `ggtree` [@Yu-2017]

@@

DSF was supported by Australian Research Council Future Fellowship (FT160100113), RVG was supported by Australian Research Council DECRA Fellowship (DE170100208). This work was supported by fellowship grants from Australian Research Council to Falster (FT160100113), Gallagher (DE170100208) and Wright, and funding from the Australian Research Data Commons under their "Transformative data collections" program.

We acknowledge XXXXXXXXXXx institutions, who XXXXX

@@

RVG, IJW conceived the original idea; RVG, EHW, CB collated data from primary sources; DSF developed the workflow for the harmonising of data and led all coding; EHW, DI, SCA, JL contributed to coding; EHW, CB, JL, SA error-checked trait observations; DI developed figures for the paper; DSF, RVG, DI, EHW wrote the first draft of the paper. All other authors contributed the raw data and metadata underpinning the resource and reviewed the harmonised data for errors.

@@

```{r austraits}
## Assumes to items exist in global name space

@@

sources_all <- 
  bind_rows(
    austraits$methods %>% select(dataset_id, citation = source_primary_key),
    austraits$methods %>% select(dataset_id, citation = source_secondary_key)
  ) %>% 
  na_if("") %>%
  na.omit() %>% 
  distinct() %>% 
  arrange(dataset_id) %>% 
  group_by(dataset_id) %>% 
  summarise(refs = sprintf("@%s", citation) %>% str_replace_all("; ", "; @") %>% paste(collapse = "; ") )

@@

@@

\newpage

@@

n_records <- 
  data_geo %>% 
  group_by(trait_name) %>%
  summarise(
            Description = "",
            Type= "",
            `Records (all)` = n(),
            `Records (geo.)` = sum(!is.na(`longitude (deg)`) & !is.na(`latitude (deg)`)),
            Studies = n_distinct(dataset_id),
            Taxon = n_distinct(taxon_name),
            Families = n_distinct(family),
            `Data Citations` = refs %>% unique() %>% paste(collapse = "; ") %>% paste0("[", ., "]")
            ) %>% 
  left_join(trait_categories, by=c("trait_name")) %>% 
  select(Tissue = tissue, Category = category, Trait = trait_name, everything()) %>% 
  arrange(Tissue, Category, Trait)

n_records$Type <- sapply(austraits$definitions$traits$elements, "[[","type")[n_records$Trait]
n_records$Description <- sapply(austraits$definitions$traits$elements, "[[","description")[n_records$Trait] %>% unlist()

n_records %>%
  select(-Tissue, -Description) %>%
  knitr::kable()
```

@@
